---
title: "QC document"
author: "Fredrik Landfors"
date: "12/2/2020"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
# Set global options
knitr::opts_chunk$set(echo = TRUE, warning = TRUE)
```

Source libraries and functions:
```{r}
source("./src/check_packages.R")
check_packages(
  cran_packages = c("ggplot2", "naniar", "UpSetR"),
  bioc_packages = c("ropls")
)
library(ggplot2)
source("./src/plot_pca.R")
source("./src/geom_cwise_boxplot.R")
source("./src/geom_rwise_boxplot.R")
```

Load data:
```{r}
load("~/projekt_data/2019-11-19_SCD-FIA3_data/rData/markdown.RData")
```

# QC: Proximity extension assay

Get data:
```{r}
qc <- list()
preanalytic_factors <- c(
  "Patient_code",
  "FIAnum_FIA3",
  "Set_FIA3",
  "Sample_date",
  "case_freeze_thawed",
  "Fast_sample",
  "Age", 
  "Sex"
)
randomization_factors <- c(
  "Glc_0h",
  "Chol_tot",
  "BMI",
  "Smoker_2fct",
  "Sbt_VIP",
  "Case_control"
)
qc$inf <- merge(
  x = preproc$raw_fia$data_out[c(preanalytic_factors, randomization_factors)],
  y = preproc$raw_inf$data_out,
  by = "Patient_code"
)
qc$org <- merge(
  x = preproc$raw_fia$data_out[c(preanalytic_factors, randomization_factors)],
  y = preproc$raw_org$data_out,
  by = "Patient_code"
)
qc$fia <- merge(
  x = preproc$raw_org$data_out["Patient_code"],
  y = preproc$raw_fia$data_out[c("Patient_code", "Case_control", imp_fia_vars)],
  by = "Patient_code"
)
qc$org_proteins <- names(preproc$raw_org$data_out)[
  names(preproc$raw_org$data_out) %in% preproc$raw_org$varmeta_data$name1
]
qc$inf_proteins <- names(preproc$raw_inf$data_out)[
  names(preproc$raw_inf$data_out) %in% preproc$raw_inf$varmeta_data$name1
]
```

## Olink ORGAN DAMAGE (v.3311)

### A. Missing values
```{r}
naniar::vis_miss(qc$org)
```

### B. Lower limit of detection (LOD)

Values with > 5 % of values below LOD were excluded from further analysis. This reduced the count of proteins by 39 from 92 to 53.
```{r}
length(qc$org_proteins)
length(preproc$raw_org$varmeta_data$name1)
length(preproc$raw_org$varmeta_data$name1) - length(qc$org_proteins)
```

### C. Batch effects

PCA colored for plate ID
```{r}
plot_pca(
      x = qc$org[, qc$org_proteins],
      c = qc$org["Plate_ID_2"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```

PLS model 
```{r}
# ropls::opls(
#   x = qc$org[, qc$org_proteins],
#   y = qc$org[, "Plate_ID_2"]
# )
```
PLS model: force two predictive components because first model did not pass permutation test
```{r}
ropls::opls(
  x = qc$org[, qc$org_proteins],
  y = qc$org[, "Plate_ID_2"],
  predI = 2
)
```

### D. Preanalytic effects

#### Freeze-thaw cycles
```{r}
plot_pca(
      x = qc$org[, qc$org_proteins],
      c = qc$org["case_freeze_thawed"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls1 <- ropls::opls(
  x = qc$org[, qc$org_proteins],
  y = qc$org[, "case_freeze_thawed"]
)
ropls::plot(pls1, typeVc = "xy-weight")
```

#### Fasting
```{r}
plot_pca(
      x = qc$org[, qc$org_proteins],
      c = qc$org["Fast_sample"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls2 <- ropls::opls(
  x = qc$org[, qc$org_proteins],
  y = qc$org[, "Fast_sample"],
  predI = 2
)
```

#### Sample date
```{r}
plot_pca(
      x = qc$org[, qc$org_proteins],
      c = qc$org["Sample_date"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls3 <- ropls::opls(
  x = qc$org[, qc$org_proteins],
  y = sapply(
    qc$org[, "Sample_date"] ,
      function(x) {
        round(((as.numeric(x) - 6159) / 365.25 + 1986.86242), 0)
      }
    )
)
ropls::plot(pls3, typeVc = "xy-weight")
```

#### Age at sampling
```{r}
plot_pca(
      x = qc$org[, qc$org_proteins],
      c = qc$org["Age"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls4 <- ropls::opls(
  x = qc$org[, qc$org_proteins],
  y = qc$org[, "Age"]
)
ropls::plot(pls4, typeVc = "xy-weight")
```

#### Sex
```{r}
plot_pca(
      x = qc$org[, qc$org_proteins],
      c = qc$org["Sex"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls4 <- ropls::opls(
  x = qc$org[, qc$org_proteins],
  y = qc$org[, "Sex"]
)
ropls::plot(pls4, typeVc = "xy-weight")
```

### E. Olink QC flag

Looking at QC-flagged samples in plates 1-5 (coloured blue):

```{r}
geom_rwise_boxplot(
      x = subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-1", select = qc$org_proteins),
      fill1 = qc$org$QC_Warning_2
)
geom_rwise_boxplot(
      x = subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-2", select = qc$org_proteins),
      fill1 = qc$org$QC_Warning_2
)
geom_rwise_boxplot(
      x = subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-3", select = qc$org_proteins),
      fill1 = qc$org$QC_Warning_2
)
geom_rwise_boxplot(
      x = subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-4", select = qc$org_proteins),
      fill1 = qc$org$QC_Warning_2
)
geom_rwise_boxplot(
      x = subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-5", select = qc$org_proteins),
      fill1 = qc$org$QC_Warning_2
)
```

### F. Plate randomization

Checks if case/ctrl pairs were analyzed on the same plates
```{r}
table(table(subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-1", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-2", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-3", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-4", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, Plate_ID_2 == "FIA3_2-ORG-5", select = "Set_FIA3")) == 2)
```

Check covariate randomization across plates
```{r}
# BMI
anova(lm(formula = BMI ~ Plate_ID_2, data = qc$org))
# Total cholesterol
anova(lm(formula = Chol_tot ~ Plate_ID_2, data = qc$org))
# Systolic blood pressure
anova(lm(formula = Sbt_VIP ~ Plate_ID_2, data = qc$org))
summary(lm(formula = Sbt_VIP ~ Plate_ID_2, data = qc$org))
# Glucose
anova(lm(formula = Glc_0h ~ Plate_ID_2, data = qc$org))
# Smoking status
table(qc$org[c("Smoker_2fct", "Plate_ID_2")])
chisq.test(table(qc$org[c("Smoker_2fct", "Plate_ID_2")]))
```

Check pre-analytic factor randomization across plates
```{r}
# Sample date
anova(lm(formula = as.numeric(Sample_date) ~ Plate_ID_2, data = qc$org))
summary(lm(formula = as.numeric(Sample_date) ~ Plate_ID_2, data = qc$org))
# Fasting
table(qc$org[c("Fast_sample", "Plate_ID_2")])
chisq.test(table(qc$org[c("Fast_sample", "Plate_ID_2")]))
table(table(subset(qc$org, Fast_sample == "0-4h", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, Fast_sample == "4-6h", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, Fast_sample == "6-8h", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, Fast_sample == ">8h", select = "Set_FIA3")) == 2)
# Freeze/thaw status
table(qc$org[c("case_freeze_thawed", "Plate_ID_2")])
chisq.test(table(qc$org[c("case_freeze_thawed", "Plate_ID_2")]))
table(table(subset(qc$org, case_freeze_thawed == "NO", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, case_freeze_thawed == "ONCE", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, case_freeze_thawed == "Thawed due to broken freezer but still cold", select = "Set_FIA3")) == 2)
table(table(subset(qc$org, case_freeze_thawed == "TWICE", select = "Set_FIA3")) == 2)
```

Case control matching was perfect for freeze/thaw cycles, but not for fasting time.

Check fasting time overlap using a UpSet plot.
```{r}
qc_split <- split(
  qc$org[c(preanalytic_factors, randomization_factors)],
  f = qc$org$Case_control
)
qc_overlap <- merge(
  x = qc_split$Ctrl[c("Set_FIA3", "Fast_sample")],
  y = qc_split$Case[c("Set_FIA3", "Fast_sample")],
  by = "Set_FIA3"
)
names(qc_overlap) <- c("Set_FIA3", "ctrl_", "case_")
qc_overlap2 <- data.frame(
  qc_overlap["Set_FIA3"],
  model.matrix(~ ctrl_ - 1, data = qc_overlap["ctrl_"]),
  model.matrix(~ case_ - 1, data = qc_overlap["case_"])
)
UpSetR::upset(
  qc_overlap2,
  #nsets = 8, 
  sets = c("ctrl_0.4h",
           "case_0.4h",
           "ctrl_4.6h",
           "case_4.6h",
           "ctrl_6.8h",
           "case_6.8h",
           "ctrl_.8h",
           "case_.8h"),
  keep.order = TRUE
)
```
Conclusion: 12 pairs where fasted subjects were matched with recently fed subjects.

## Olink INFLAMMATION (v.3021)

### A. Missing values
```{r}
naniar::vis_miss(qc$inf)
```

### B. Lower limit of detection (LOD)

Values with > 5 % of values below LOD were excluded from further analysis. This reduced the count of proteins by 23 from 92 to 69.
```{r}
length(qc$inf_proteins)
length(preproc$raw_inf$varmeta_data$name1)
length(preproc$raw_inf$varmeta_data$name1) - length(qc$inf_proteins)
```

### C. Batch effects

PCA colored for plate ID
```{r}
plot_pca(
      x = qc$inf[, qc$inf_proteins],
      c = qc$inf["Plate_ID_1"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```

PLS model 
```{r}
# ropls::opls(
#   x = qc$inf[, qc$inf_proteins],
#   y = qc$inf[, "Plate_ID_1"]
# )
```
PLS model: force two predictive components
```{r}
ropls::opls(
  x = qc$inf[, qc$inf_proteins],
  y = qc$inf[, "Plate_ID_1"],
  predI = 2
)
```

### D. Preanalytic effects

#### Freeze-thaw cycles
```{r}
plot_pca(
      x = qc$inf[, qc$inf_proteins],
      c = qc$inf["case_freeze_thawed"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls1 <- ropls::opls(
  x = qc$inf[, qc$inf_proteins],
  y = qc$inf[, "case_freeze_thawed"]
)
ropls::plot(pls1, typeVc = "xy-weight")
```

#### Fasting
```{r}
plot_pca(
      x = qc$inf[, qc$inf_proteins],
      c = qc$inf["Fast_sample"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls2 <- ropls::opls(
  x = qc$inf[, qc$inf_proteins],
  y = qc$inf[, "Fast_sample"],
  predI = 2
)
```

#### Sample date
```{r}
plot_pca(
      x = qc$inf[, qc$inf_proteins],
      c = qc$inf["Sample_date"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls3 <- ropls::opls(
  x = qc$inf[, qc$inf_proteins],
  y = sapply(
    qc$inf[, "Sample_date"] ,
      function(x) {
        round(((as.numeric(x) - 6159) / 365.25 + 1986.86242), 0)
      }
    )
)
ropls::plot(pls3, typeVc = "xy-weight")
```

#### Age at sampling
```{r}
plot_pca(
      x = qc$inf[, qc$inf_proteins],
      c = qc$inf["Age"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls4 <- ropls::opls(
  x = qc$inf[, qc$inf_proteins],
  y = qc$inf[, "Age"]
)
ropls::plot(pls4, typeVc = "xy-weight")
```

#### Sex
```{r}
plot_pca(
      x = qc$inf[, qc$inf_proteins],
      c = qc$inf["Sex"],
      opt_ellipse = "auto",
      return_fit = FALSE,
      n_pc = 7
)
```
```{r}
pls4 <- ropls::opls(
  x = qc$inf[, qc$inf_proteins],
  y = qc$inf[, "Sex"]
)
ropls::plot(pls4, typeVc = "xy-weight")
```

### E. Olink QC flag

Looking at QC-flagged samples in plates 1-5 (coloured blue):

```{r}
geom_rwise_boxplot(
      x = subset(qc$inf, Plate_ID_1 == "FIA3_2-INF-1", select = qc$inf_proteins),
      fill1 = qc$inf$QC_Warning_1
)
geom_rwise_boxplot(
      x = subset(qc$inf, Plate_ID_1 == "FIA3_2-INF-2", select = qc$inf_proteins),
      fill1 = qc$inf$QC_Warning_1
)
geom_rwise_boxplot(
      x = subset(qc$inf, Plate_ID_1 == "FIA3_2-INF-3", select = qc$inf_proteins),
      fill1 = qc$inf$QC_Warning_1
)
geom_rwise_boxplot(
      x = subset(qc$inf, Plate_ID_1 == "FIA3_2-INF-4", select = qc$inf_proteins),
      fill1 = qc$inf$QC_Warning_1
)
geom_rwise_boxplot(
      x = subset(qc$inf, Plate_ID_1 == "FIA3_2-INF5", select = qc$inf_proteins),
      fill1 = qc$inf$QC_Warning_1
)
```

### F. Plate randomization

Checks if case/ctrl pairs were analyzed on the same plates
```{r}
table(table(subset(qc$inf, Plate_ID_1 == "FIA3_2-INF-1", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, Plate_ID_1 == "FIA3_2-INF-2", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, Plate_ID_1 == "FIA3_2-INF-3", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, Plate_ID_1 == "FIA3_2-INF-4", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, Plate_ID_1 == "FIA3_2-INF5", select = "Set_FIA3")) == 2)
```

Check covariate randomization across plates
```{r}
# BMI
anova(lm(formula = BMI ~ Plate_ID_1, data = qc$inf))
# Total cholesterol
anova(lm(formula = Chol_tot ~ Plate_ID_1, data = qc$inf))
# Systolic blood pressure
anova(lm(formula = Sbt_VIP ~ Plate_ID_1, data = qc$inf))
summary(lm(formula = Sbt_VIP ~ Plate_ID_1, data = qc$inf))
# Glucose
anova(lm(formula = Glc_0h ~ Plate_ID_1, data = qc$inf))
# Smoking status
table(qc$inf[c("Smoker_2fct", "Plate_ID_1")])
chisq.test(table(qc$inf[c("Smoker_2fct", "Plate_ID_1")]))
```

Check pre-analytic factor randomization across plates
```{r}
# Sample date
anova(lm(formula = as.numeric(Sample_date) ~ Plate_ID_1, data = qc$inf))
summary(lm(formula = as.numeric(Sample_date) ~ Plate_ID_1, data = qc$inf))
# Fasting
table(qc$inf[c("Fast_sample", "Plate_ID_1")])
chisq.test(table(qc$inf[c("Fast_sample", "Plate_ID_1")]))
table(table(subset(qc$inf, Fast_sample == "0-4h", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, Fast_sample == "4-6h", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, Fast_sample == "6-8h", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, Fast_sample == ">8h", select = "Set_FIA3")) == 2)
# Freeze/thaw status
table(qc$inf[c("case_freeze_thawed", "Plate_ID_1")])
chisq.test(table(qc$inf[c("case_freeze_thawed", "Plate_ID_1")]))
table(table(subset(qc$inf, case_freeze_thawed == "NO", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, case_freeze_thawed == "ONCE", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, case_freeze_thawed == "Thawed due to broken freezer but still cold", select = "Set_FIA3")) == 2)
table(table(subset(qc$inf, case_freeze_thawed == "TWICE", select = "Set_FIA3")) == 2)
```

Case control matching was perfect for freeze/thaw cycles, but not for fasting time.

Check fasting time overlap using a UpSet plot.
```{r}
qc_split <- split(
  qc$inf[c(preanalytic_factors, randomization_factors)],
  f = qc$inf$Case_control
)
qc_overlap <- merge(
  x = qc_split$Ctrl[c("Set_FIA3", "Fast_sample")],
  y = qc_split$Case[c("Set_FIA3", "Fast_sample")],
  by = "Set_FIA3"
)
names(qc_overlap) <- c("Set_FIA3", "ctrl_", "case_")
qc_overlap2 <- data.frame(
  qc_overlap["Set_FIA3"],
  model.matrix(~ ctrl_ - 1, data = qc_overlap["ctrl_"]),
  model.matrix(~ case_ - 1, data = qc_overlap["case_"])
)
UpSetR::upset(
  qc_overlap2,
  #nsets = 8, 
  sets = c("ctrl_0.4h",
           "case_0.4h",
           "ctrl_4.6h",
           "case_4.6h",
           "ctrl_6.8h",
           "case_6.8h",
           "ctrl_.8h",
           "case_.8h"),
  keep.order = TRUE
)
```
Conclusion: 12 pairs where fasted subjects were matched with recently fed subjects.

# QC: FIA missing values

## A. Pre-imputation diagnostics

Percent missing per variable (n = 448):

```{r}
preproc$mice_fia$functions$show_missing(
  qc$fia
)
naniar::vis_miss(qc$fia)
naniar::vis_miss(subset(qc$fia, Case_control == "Ctrl"))
naniar::vis_miss(subset(qc$fia, Case_control == "Case"))
```

OBS! The whole FIA data set (n = 3638) is used for imputation of clinical
variables. Missing values are imputed using MICE.

Percent missing per variable (n = 3638):

```{r}
dim(preproc$raw_fia$data_out)
preproc$mice_fia$data$percent_missing
preproc$mice_fia$plots$vismissplot
```

Fluxplot shows if imputation for selected variables is feasible:

```{r}
mice::fluxplot(preproc$raw_fia$data_out[imp_fia_vars])
```

## B. Post-imputation diagnostics 

Density and stripplots post-imputation shows imputed variables were within
reasonable ranges:

```{r}
preproc$mice_fia$plots$mice_density
preproc$mice_fia$plots$mice_stripplot
```

## C. Outlier check

```{r}
boxplot(
  sapply(
    Filter(is.numeric, preproc$imp_fia),
    scale
  )
)
```